Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-08-08 submission/manuscript.Rmd
add motif results paragraph

- read in the homer results txt file above here
- Add in euler KDM5C-bound vs KDM5C-unbound coloring
		- Make sure instances window is 500 bp, change if not

- Germline genes repressed by other histone modifying enzymes have shared genomic features around the TSS to suppress their expression, such as motifs for E2F6/DP1 (E2F) and MGA/MAX (E-box) and CpG islands (CGIs). We asessed transcription factor motifs enriched at germline gene promoters that are either bound or unbound by KDM5C in EpiLCs using HOMER[XXX] (TSS +/- 500 bp, q-value < 0.1). MAX and E2F6 binding sites were significantly enriched at KDM5C-bound germline genes, with about one third of KDM5C-bound promoters contained one or both of these motifs (Figure 4). Contrastingly, KDM5C-unbound germline genes were enriched for RFX transcription factor motifs, a family of transcription factors important for ciliogenesis. The RFX X-box motif is present at XXX% of KDM5C-unbound promoters (Figure 4).

	- Enriched RFX transcription factors included the spermiogenesis regulator RFX2, whose mRNA is also significantly upregulated in _Kdm5c_-KO EpiLCs (Figure 4). 
	 - https://www.nature.com/articles/srep20435
	 - https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4498915/
	 - https://pubmed.ncbi.nlm.nih.gov/32725242/

@@

- Germline genes are known to accumulate CpGme at (CGIs).
To determine KDM5C's role in the initial placement of DNA methylation at germline gene promoters, we perfomed whole genome bislufite sequencing (WGBS) in wild-type and _Kdm5c_-KO ESCs and 96 hour extended EpiLCs (exEpiLCs). Although wild-type cells accumulated high levels of DNA methylation at germline gene promoters over the course of ESCs to exEpiLC differentiation, DNA methylation was markedly reduced in _Kdm5c_-KO exEpiLCs (Figure 5F).

@@

Motif Discussion
- only 1/3 had max/e2f - Unlikely to be the major recruitment mechanism for KDM5C (discussion?)

@@

>>>>

Many germline-specific genes, including meiotic genes, are suppressed by the transcription factor heterodimers E2F6/DP1 and MGA/MAX, which respectively bind E2F and E-box motifs[@hurlinMgaDualspecificityTranscription1999; @pohlersRoleE2F6Restriction2005; @velascoDnmt3bRecruitmentE2F62010; @stielowMGAL3MBTL2E2F62018; @dahletE2F6InitiatesStable2021]. Thus, we identified transcription factor motifs enriched at KDM5C-bound or unbound germline gene promoters using HOMER[@heinzSimpleCombinationsLineageDetermining2010] (TSS +/- 500 bp, q-value < 0.1). MAX and E2F6 binding sites were significantly enriched at germline genes bound by KDM5C in EpiLCs, but not at germline genes unbound by KDM5C (MAX q-value: , E2F6 q-value:, E2F q-value: ) (Figure 4). One third of KDM5C-bound promoters contained the consensus sequence for either E2F6 (E2F, 5'-TCCCGC-3'), MGA (E-box, 5'-CACGTG-3'), or both, but only 17% of KDM5C-unbound genes contained these motifs (Figure 4). KDM5C-unbound germline genes were intstead enriched for multiple RFX transcription factor binding sites. RFX transcription factors bind X-box motifs[@gajiwalaStructureWingedhelixProtein2000] to promote ciliogenesis[@swobodaRFXTypeTranscriptionFactor2000; @ashiqueRfx4TranscriptionFactor2009]. Enriched RFX trancription factors included RFX2, a central regulator of post-meiotic spermatogeneis[@kistlerRFX2MajorTranscriptional2015; @wuTranscriptionFactorRFX22016]. Interestingly, RFX2 mRNA is upregualted in _Kdm5c_-KO EpiLCs, but is also not a direct target of KDM5C (Figure 4). Thus, distinct transcription factor programs regulate KDM5C-bound and unbound germline genes.

@@

- Germline genes are known to accumulate CpGme at (CGIs).
	- What percentage of germline genes have CGIs
- Do differential methylation analysis for WT ESCs to WT EpiLCs
	- What percentage of germline genes significanty gain methylation (at CGI or at promoter)
	- Out of the ones that gain methylation, which are significantly reduced

To determine KDM5C's role in the initial placement of DNA methylation at germline gene promoters, we perfomed whole genome bislufite sequencing (WGBS) in wild-type and _Kdm5c_-KO ESCs and 96 hour extended EpiLCs (exEpiLCs). Although wild-type cells accumulated high levels of DNA methylation at germline gene promoters over the course of ESCs to exEpiLC differentiation, DNA methylation was markedly reduced in _Kdm5c_-KO exEpiLCs (Figure 5F).

@@

@@

- Motif analysis
	- only 1/3 had max/e2f - Unlikely to be the major recruitment mechanism for KDM5C
	- KDM5C does not contain motif-specific binding
	- controls cilia in mutiple cell types, initially identified in neurons. Shared axis that could contribute to NDDs
		- RFX Cilia in NDDs - https://pubmed.ncbi.nlm.nih.gov/33658631/ 
	- Enriched RFX transcription factors included the spermiogenesis regulator RFX2, whose mRNA is also significantly upregulated in _Kdm5c_-KO EpiLCs (Figure 4). 
	 - https://www.nature.com/articles/srep20435
	 - https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4498915/
	 - https://pubmed.ncbi.nlm.nih.gov/32725242/
